---
title: Annotated equations in ggplot2
date: '2023-07-24'
execute: 
  cache: true
categories: 
  - ggplot2
description: "Importing latex into ggplot2"
tags:
  - ggplot2
  - R
---

```{r setup}
#| include: false
#| echo: false

options(tidyverse.quiet = TRUE)
library(conflicted)
conflicts_prefer(dplyr::filter, 
                 dplyr::lag,
                 .quiet = TRUE)
library(tidyverse)
library(ggtext)
library(ggimage)
library(WJSmisc)
library(showtext)
font_add_google("Roboto Condensed", "Roboto Condensed")
showtext_auto()


```


:::{.callout-note}
The [xdvir package](https://www.stat.auckland.ac.nz/~paul/Reports/dvir/xdvir/xdvir.html), which inserts $\LaTeX$ equations into grid-based plots (including ggplot2) is now on CRAN. I recommend xdvir over the older methods presented in this post. Other than this message, I have left this post "as is."
:::

R has a mathematical annotation system via [plotmath](https://stat.ethz.ch/R-manual/R-devel/library/grDevices/html/plotmath.html), but I like the look of true $\LaTeX$ equations better. 

Getting $\LaTeX$ equations into ggplot2 plots has never been easy. The [tikzdevice](https://cran.r-project.org/package=tikzDevice) package is great if you are generating a .pdf document. If you are not, then you might want to consider other options.

The easiest, hassle-free option that I know of is to create the equation in a $\TeX$ editor and then import the resulting .pdf using [ggimage](https://github.com/GuangchuangYu/ggimage). If I need the best possible image quality, I convert the .pdf to .svg using `dvisvgm` (see example below).

For annotated equations, I like using the aptly-named [annotate-equations](https://ctan.org/pkg/annotate-equations?lang=en) $\LaTeX$ package. It uses tikz to remember where parts of equations are on the page. The `\eqnmark` and `\eqnmarkbox` functions work like so:

`\eqnmark[color]{node_name}{latex equation terms}`

Then use the `\annotate` function like so:

`\annotate[color]{above,left}{node_name}{annotation text}`

# My Annotated Equation

I am going to refer to the same file name with various endings (e.g., .tex, .pdf, .svg), so I will define it here.

```{r myfile}
myfile <- "annotatedequationsimple"
```


Here is the code for the annotated equation. It is saved in `annotatedequationsimple.tex`.

```{cat zscore, engine.opts=list(file=paste0(myfile, ".tex"))}
\documentclass[border={25pt 50pt -35pt 52pt}]{standalone}
%\documentclass{article}
\usepackage{annotate-equations}
\usepackage{xcolor}
\definecolor{myviolet}{HTML}{440154}
\definecolor{myblue}{HTML}{3B528B}
\definecolor{myindigo}{HTML}{21908C}
\definecolor{mygreen}{HTML}{5DC863}
\usepackage[sfdefault,condensed]{roboto}
\begin{document}
	
\renewcommand{\eqnhighlightheight}{\mathstrut}
	
\huge$
\eqnmark[myviolet]{z}{z} = 
\frac{
	\eqnmark[myblue]{x}{X}-
	\eqnmark[myindigo]{mu}{\mu}}{
	\eqnmark[mygreen]{sigma}{\sigma}}
$


\annotate[
	yshift=1em, 
	myviolet,
	align=right]
	{above, left}
	{z}
	{$z$-score}
	
\annotate[
	yshift=1em, 
	myblue, 
	align=right]
	{above,left}
	{x}
	{Observed\\ Score}
	

\annotate[
	yshift=1em, 
	myindigo]
	{above,right}
	{mu}
	{Population\\ Mean\\ $\mu = 100$}
	

	
\annotate[
	yshift=-.4em, 
    mygreen, 
    align=right]
    {below,left}
    {sigma}
    {Population\\ Standard\\ Deviation\\ $\sigma = 15$}
	
\end{document}

```

```{tikz zscorecode}
#| code-fold: show
#| code-summary: "LaTeX code"
#| echo: fenced
#| eval: false

<<zscore>>
```

Now convert the .tex file to .pdf:

```{r pdfconvert}
#| results: hide
paste0('pdflatex -interaction=nonstopmode ', myfile,'.tex') |> 
  shell()
```

Now we can convert the .pdf to .svg:

```{r svgconvert}
paste0("dvisvgm --pdf --output=", myfile,".svg ", myfile,".pdf") |> 
  shell()
```

For best image quality, import the .svg file with [svgparser](https://coolbutuseless.github.io/package/svgparser/index.html). The `read_svg` function will create a grid grob that can plotted directly using `ggplot2::annotation_custom`. [If you want fine control of the imported grob, you can plot it using the [gggrid](https://www.stat.auckland.ac.nz/~paul/Reports/gggrid/gggrid.html#gggrid) package. See [here](https://coolbutuseless.github.io/package/svgparser/articles/gggrid.html) for an example.]{.column-margin}

In the simplest case, we can do this:

```{r fig-simpleplot, dev='svglite', fig.showtext=TRUE}
#| fig-cap: A simple plot with an annotated equation.
library(svgparser)
my_svg <- svgparser::read_svg(paste0(myfile, ".svg"))
ggplot() +
  theme_void() +
  annotation_custom(my_svg, xmin = 0, xmax = 1, ymin = 0, ymax = 1) 
```

However, this is no better than just displaying the .svg directly. You probably want to embed the equation in a plot. For example:

```{r fig-svg, dev='svglite', fig.showtext=TRUE}
#| out-width: 100%
#| fig-cap: Importing an .svg file via svgparser and `annotation_custom`.
#| code-fold: true

mu <- 100
sigma <- 15
plot_height <- dnorm(mu, mu, sigma)
lb <- -4 * sigma + mu
ub <- 4 * sigma + mu

ggplot() +
  annotation_custom(my_svg,
                    xmin = 112,
                    xmax = 164,
                    ymin = .33 * plot_height) +
  stat_function(
    fun = \(x) dnorm(x, mean = mu, sd = sigma),
    geom = "area",
    n = 1000,
    fill = "dodgerblue",
    alpha = .5
  ) +
  theme_classic(base_family = "Roboto Condensed",
                base_size = 18) +
  theme(
    axis.text.x = element_markdown(),
    axis.title.x = element_markdown(),
    axis.line = element_blank()
  )  +
  scale_x_continuous(
    "Observed Score *X*<br>*z*",
    breaks = seq(lb, ub, sigma),
    limits = c(lb, ub),
    labels = \(x) paste0(
      signs::signs(x),
      "<br>",
      ifelse(
        x == mu,
        "<em>&mu;</em>",
        paste0(
          signs::signs((x - mu) / sigma,
                       add_plusses = T,
                       label_at_zero = "none"
          ),
          "<em>&sigma;</em>"
        )
      )
    )
  ) +
  scale_y_continuous(
    NULL,
    limits = c(0, plot_height),
    expand = expansion(),
    breaks = NULL
  ) 
```


If you can live with just a little pixelation, the [ggimage](https://github.com/GuangchuangYu/ggimage) package can import a .pdf directly with good results and less hassle, provided you render the plot with the [ragg](https://ragg.r-lib.org/) package.

```{r fig-pdf, dev='ragg_png', dpi=300, fig.showtext=TRUE}
#| fig-cap: Importing a .pdf file via `ggimage::geom_image`.
#| out-width: 100%
#| code-fold: true
ggplot() +
  geom_image(
    data = tibble(
      x = 140,
      y = .65 * plot_height,
      image = "annotatedequationsimple.pdf"
    ),
    aes(x, y, image = image),
    size = .70
  ) +
  stat_function(
    fun = \(x) dnorm(x, mean = mu, sd = sigma),
    geom = "area",
    n = 1000,
    fill = "dodgerblue",
    alpha = .5
  ) +
  theme_classic(base_family = "Roboto Condensed", 
                base_size = 18) +
  theme(
    axis.text.x = element_markdown(),
    axis.title.x = element_markdown(),
    axis.line = element_blank()
  )  +
  scale_x_continuous(
    "Observed Score *X*<br>*z*",
    breaks = seq(lb, ub, sigma),
    limits = c(lb, ub),
    labels = \(x) paste0(
      signs::signs(x),
      "<br>",
      ifelse(
        x == mu,
        "<em>&mu;</em>",
        paste0(
          signs::signs((x - mu) / sigma,
                       add_plusses = T,
                       label_at_zero = "none"
          ),
          "<em>&sigma;</em>"
        )
      )
    )
  ) +
  scale_y_continuous(
    NULL,
    limits = c(0, plot_height),
    expand = expansion(),
    breaks = NULL
  )
```


# A more complex example

In this example, I used the `\eqnmarkbox` function for greater clarity. The .tex file is saved in a file called `annotatedequation.tex`.

```{tikz annotatedequation}
#| eval: false
#| code-fold: true
#| code-summary: "LaTeX code"

\documentclass[border={10pt 48pt -45pt 62pt}]{standalone}
%\documentclass{article}
\usepackage{annotate-equations}
\usepackage{xcolor}
\definecolor{myviolet}{HTML}{414487}
\definecolor{myblue}{HTML}{2F6C8E}
\definecolor{myblue2}{HTML}{21908C}
\definecolor{mygreen}{HTML}{2FB47C}
\definecolor{mygreen2}{HTML}{7AD151}
\usepackage[sfdefault,condensed]{roboto}
\begin{document}
	
\renewcommand{\eqnhighlightheight}{\mathstrut}
	
$\LARGE
\eqnmarkbox[myviolet]{nodeP}{P\left(T \le \tau \right)} = 
\eqnmarkbox[myblue]{phi}{\Phi}
\left(\frac{
	\eqnmarkbox[myblue2]{tau}{\tau}-
	\eqnmarkbox[mygreen]{esttrue}{\hat{T}}}{
	\eqnmarkbox[mygreen2]{sigma}{\sigma_{T - \hat{T}}}}
\right)$


\annotate[
	yshift=1em, 
	xshift=11mm, 
	myviolet]
	{above, left}
	{nodeP}
	{Probability true score $T$\\ is less than threshold $\tau$ }
	
\annotate[
	yshift=-.6em,
	myblue]
	{below,left}
	{phi}
	{Standard Normal Cumulative\\ Distribution Function $\Phi()$}
	
\annotate[
	yshift=1.4em, 
	xshift=4mm, 
	myblue2]
	{above,left}
	{tau}
	{Threshold $\tau=70$}
	
\annotate[
	yshift=3em, 
	xshift=7mm, 
	mygreen]
    {above,left}
    {esttrue}
    {Estimated True Score $\hat{T}=r_{XX}(X-\mu)+\mu$\\ 
    Observed Score $X\sim \mathcal{N}\left(\mu = 100, \sigma=15\right)$\\ 
    Reliability Coefficient $r_{XX}=\{.80,.85,.90,.95,.98\}$}
	
\annotate[
	yshift=-2em, 
    mygreen2]
    {below,left}
    {sigma}
    {Standard Error of the Estimate\\ 
    $\sigma_{T-\hat{T}}=\sigma\sqrt{r_{XX}-r_{XX}^2}$}
	
\end{document}
```

Now convert .tex to .pdf:

```{r pdfconvert2}
#| results: hide
#| code-fold: true

myfile <- "annotatedequation"

paste0('pdflatex -interaction=nonstopmode ', myfile,'.tex') |> 
  shell()
```

And we are ready to plot. This plot shows the probability that an observed score will have a true score less than a specific threshold, given a reliability coefficient.

```{r fig-pthresholdequation, fig.showtext=TRUE}
#| fig-cap: 'The probability that a true score is less than 70 depends on the observed score’s position and reliability coefficient.'
#| out-width: 100%
#| fig-width: 6.5
#| dev: ragg_png
#| code-fold: true
#| classes: preview-image


viridis_start <- .2
viridis_end <- .8
threshold <- 70

# Find where a line intersects with the normal cdf
find_x <- function(rxx = .8,
                   slope = 0.0048,
                   intercept = .66,
                   mu = 100,
                   sigma = 15,
                   start_x = 60,
                   threshold = 70) {
  x <- start_x
  precision <- .00001
  diff_y <- precision * 10
  reps <- 0
  while (abs(diff_y) > precision) {
    x <- x - diff_y 
    line_y <- x * slope + intercept
    curve_y <- pnorm(threshold,
                     mean = rxx * (x - mu) + mu,
                     sd = sigma * sqrt(rxx - rxx ^ 2))
    
    line_y <- x * slope + intercept
    curve_y <- pnorm(threshold,
                     mean = rxx * (x - mu) + mu,
                     sd = sigma * sqrt(rxx - rxx ^ 2))
    diff_y <- line_y - curve_y
    reps <- reps + 1
  }
  tibble(x = x, p = line_y, mu = mu, sigma = sigma, threshold = threshold, reps = reps)
}




dimage = data.frame(x = 115, 
                    y = .62, 
                    image = paste0(myfile, ".pdf"))
v_rxx <- round(c(seq(0.80, 0.95, 0.05), 0.98), 2)

d_threshold <-
  crossing(x = round(seq(40, 160, 0.1), 1),
           rxx = v_rxx,
           threshold = 70) %>%
  mutate(
    see = 15 * sqrt(rxx - rxx ^ 2),
    mu = (x - 100) * rxx + 100,
    p = pnorm(threshold, mu, see)
  ) %>%
  group_by(rxx) %>%
  mutate(acceleration = p - lag(p)) %>%
  ungroup



d_labels <- tibble(rxx = v_rxx) %>%
  mutate(x = map_df(rxx, find_x)) |> 
  unnest(x)

d_threshold %>%
  ggplot(aes(x, p)) +
  geom_line(aes(color = factor(rxx)), lwd = 1) +
  geom_vline(
    aes(xintercept = threshold),
    lty = 2,
    lwd = 1,
    color = "gray30"
  ) +
  geom_image(data = dimage,
             aes(x = x,
                 y = y,
                 image = image),
             size = .87) +
  geom_richtext(
    aes(label = rxx_label,
        color = factor(rxx)),
    data = d_labels %>%
      mutate(rxx_label = prob_label(rxx)),
    angle = -67,
    size = WJSmisc::ggtext_size(13),
    label.colour = NA,
    fill = "#FFFFFF",
    family = "Roboto Condensed",
    label.margin = unit(0, "mm"),
    label.r = unit(2, "mm"),
    label.padding = unit(c(0, 0.75, 0, .5), "mm")
  ) +
  scale_x_continuous(
    "Observed Score",
    breaks = seq(40, 160, 15),
    minor_breaks = seq(40, 160, 5),
    expand = expansion()
  ) +
  scale_y_continuous(
    paste0("Probability True Score < ", threshold),
    expand = expansion(),
    breaks = seq(0, 1, 0.1),
    labels = prob_label,
    limits = c(0, 1)
  ) +
  scale_color_viridis_d(begin = viridis_start,
                        end = viridis_end) +
  theme_minimal(base_family = "Roboto Condensed", 
                base_size = 16) +
  theme(legend.position = "none", 
        plot.margin = unit(c(3, 5, 2, 2), "mm")) +
  coord_fixed(ratio = 100,
              clip = "off",
              xlim = c(40, 160)) 


```